Manuscript

### 2025-04-11 Manuscript_MD_ScoreReLiability.qmd
update manuscript - add bibliography

Large-scale attempts of direct replications, using identical protocols,
such as the Many Labs studies or Registered Replication Reports [e.g.
@ManyLabs1; @RRHart], for the first time allow researchers to estimate
heterogeneity undistorted by selection processes like publication bias.
In re-analyses of these studies, Olsson-Collentine et al. [-@olsson2020]
identify a positive correlation between a phenomenon’s effect size and
its heterogeneity. Similarly, van Erp et al. [-@vanErp2017] and Stanley
et al. [-@stanley2018] estimate strong degrees of heterogeneity across
conceptual replications in Psychology. In a separate re-analysis of
large scale direct replications, Renkewitz et al. (in preparation)
identify substantial heterogeneity in almost all projects where a
non-zero effect could be identified. This aligns with the correlation
found by Olsson-Collentine et al. [-@olsson2020].

@@

The ManyLabs projects were collaborative efforts to replicate several
psychological phenomena across different research sites, employing
identical protocols. From five published projects, the data for the
first three and the fifth Many Labs projects was harmonized and publicly
stored in a standardized format when the DRIPHT repository was set up
(references). The Registered Replication Reports are similar
collaborative efforts, with the sole distinction that for each report a
single phenomenon was replicated several times. The Registered Reports
3-10 were added to the DRIPHT repository, as they employed experimental
designs with at least two groups. Lastly, some projects from the
Psychological Science Accelerator were added to the DRIPHT repository.
In contrast to Many Labs or Registered Replication Projects, these
collaborative efforts, while also distributed across the globe, do not
formally attempt to perform direct replications, but focus on original
research or conceptual replications. However, as each project used the
same protocol and participants are distributed across different
countries, a data structure similar to that of a multi-site replication
study emerges. What makes all these collaborative efforts valuable for
this manuscript is the fact that all phenomena collected in the
DRIPHT-repository are making use of two-group designs and are therefore
easily assessed using standardised mean differences.

@@

For each replication of a phenomenon, Cohen’s $d$ is computed as an
estimate of raw ES, according to @eq-d, including its standard error
(@eq-SEd). Subsequently, a random-effects meta-analysis is performed
using metafor version X.XX. To estimate meta-analytic mean ES and
heterogeneity we make ues of the REML-estimator (ref). To identify which
phenomena pass criterion (3), using a Wald-type significance test we
assess for which phenomenon the meta-analytic mean ES is significantly
different from zero. For this hypothesis test, as all other hypothesis
tests in this manuscript, we employ a one-tailed significance level of
.05.

Additionally, estimates of score reliability are derived. While not
without criticism (references), Cronbach’s Alpha is used to estimate
score reliability. It is often noted that tau-equivalence is an
unrealistic assumption to hold against real-world data, implying that
other estimates of score reliability, such as McDonald’s Omega
(reference), Guttman’s Lambda 2 or 4 (reference), or the Greatest Lower
Bound (reference) might be better suited. However, all scales employed
in this project are scored by computing the simple mean or sum of
responses across items for each individual. Computing the mean or sum of
several items implicitly assumes tau-equivalence, as each item
contributes equally to the individual’s test score. Therefore,
Cronbach’s Alpha is the better choice to estimate score reliability, as
it avoids a mismatch in assumptions between how the score is computed
and how score reliability is estimated. The individual estimates of
Cohen’s d are corrected using these estimates of Cronbach's Alpha,
according to @eq-dc, including the corrected standard errors (@eq-SEdc).
Thereby, estimates of ES are computed, which are corrected for imperfect
reliability. Again, a random-effects meta-analysis is run using metafor,
generating an estimate of heterogeneity in corrected ES. Different
indicators of heterogeneity are readily available, such as $I^2$, $H^2$
or the coefficient of variation. However, the descriptions found in H&S
and W&D discuss the absolute amount of heterogeneity in terms of
variance ($\tau^2$) or standard deviation ($\tau$). These parameters
discuss the variability of the standardized ES in the population in
terms of variance or standard deviation. Therefore, we also make use of
absolute heterogeneity here.

Subsequently, in order to assess whether ES heterogeneity was indeed
reduced by the attenuation correction procedure, the estimates of
heterogeneity in uncorrected ES and corrected ES are compared. Using
Cochran’s *Q*-test, we assess whether the estimates of heterogeneity are
statistically significantly different from 0. Only for those projects
where we have sufficient confidence that the individual estimates of
heterogeneity are larger than zero can we actually begin to interpret
whether there was any change.

To identify whether a reduction in ES heterogeneity is indeed
accompanied by differences in score reliability, the estimates of score
reliability are assessed meta-analytically as well. To do so, a
Reliability Generalization Meta-Analysis is performed (references). This
entails that individual estimates of score reliability are adequately
transformed using Bonett’s transformation:
$T_{\hat{\rho}} = ln(1 - \hat{\rho})$. This transformation has
variance-stabilising properties and therefore allows for adequate
inferences concerning heterogeneity in score reliability (references).
Cochran’s Q-test is used to identify statistically significant
heterogeneity in transformed score reliability. Estimates derived from a
Reliability Generalization Meta-Analysis using these transformations can
be back-transformed to the original score reliability scale
(references):

@@

We executed these steps for each phenomenon. The statistical programming
language R, Version X.XXX (reference) is used for all data manipulation
and statistical analysis.

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($\delta_0$) and corrected ES
($\delta$), which are central to assess the research questions posed in
section XXX. Most importantly, @tbl-tauES demonstrates that for all but
one of the phenomena, the attenuation correction procedure barely has an
impact on the extent of ES heterogeneity. However, while the
heterogeneity for the Srull_Behaviour_Hostility phenomenon seems reduced
by about a third of its initial value, the ES heterogeneity for this
phenomenon is rather small to begin with. Using a significance test, we
can not distinguish this value confidently from zero. This implies that
for the majority of phenomena, differences in score reliability can not
be made responsible for heterogeneity in standardized effect sizes.

@@

In the re-analysis of archival data in the previous section we
demonstrated that, at least for the multi-site direct replication and
PSA-projects, that score reliability does not hold any noteworthy value
as a potential moderator explaining ES heterogeneity. Additionally, we
demonstrated that the current consensus formulated in the literature
regarding how score reliability increases ES heterogeneity is false. We
found a large number of phenomena where, contrary to popular claims,
correcting for score reliability increases ES heterogeneity, instead of
decreasing it.

@@

It might help to think about linear models, when discussing the
intuition that might underlie the idea, that controlling for score
reliability should reduce ES heterogeneity. Heterogeneity describes
unexplained variation in a parameter's true distribution, that we model
using meta-analytical techniques. In that way, the idea of heterogeneity
is closely related to residual variance, as we observe it in a simple
linear model. Both describe variation that could not be explained by the
predictive variables in the case of the simple linear model, or not
explained by incorporating the Standard Errors in the case of a
meta-analytic model.

@@

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs.\@eq-HS implies that heterogeneity in $\delta_0$ is
essentially a function in which heterogeneity and mean value of $\delta$
are weighted by mean score reliability $\rho_{xx‘}$ and its
heterogeneity.

@@

We adjusted @eq-HS using the notation used throughout this text. Here,
$\delta_0$, carrying a 0 in the index, refers to the uncorrected ES,
undistorted by sampling error but not corrected for measuring error. On
the other hand $\delta$, no index, refers to the corrected ES,
undistorted by sampling error and measurement error. Lastly,
$\rho_{XX'}$ refers to score reliability. Hunter & Schmidt (xxx) claim
that @eq-HS demonstrates that differences in score reliability inflate
ES heterogeneity, which means that an attenuation correction procedure
would have to reduce ES heterogeneity. However, it can be easily
verified that @eq-HS does not allow for such claims. Depending on both
the mean value $\mu_{\rho_{xx‘}}$ and the heterogeneity
$\tau^2_{\rho_{xx‘}}$ in score reliability, ES heterogeneity might be
increased or decreased by the differences in score reliability.

Most likely, Hunter and Schmidt did not fully explore the implications
of @eq-HS, but followed the general intuition we lined out–"If a source
of variation is controlled for, the amount of unexplained variance
should be reduced." Wiernik and Dahlke (ref) refer to the work by Hunter
& Schmidt (ref), likely following their intuition.

In the re-analysis of archival data, we have demonstrated that this
intution does not hold for the relationship of ES heterogeneity and
score reliability. Instead, we will propose an alternative way that
helps us discern, under which conditions we can expect an attenuation
correction procedure to lead to an increase in ES heterogeneity.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for a different analytical description of how differences in score
reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$ (reference). First order taylor approximation may be
used to generate estimates of mean and variance of the ratio variable
\[irrespective of distribution?\] (reference). Here we use the
Taylor-estimator of the ratio’s variance to demonstrate how differences
in score reliability may affect heterogeneity in ES.

@@

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially described the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_short states the following regarding the
change in mean value and heterogeneity due to the correction: ES
heterogeneity will always be reduced, as long as the remaining
heterogeneity in score variance is not substantially smaller than the
size of the remaining mean value.

@@

It is important to note that the condition in @eq-ineq_short only
describes under which conditions the inequality is guaranteed to hold,
no matter what value other parameters like the mean value of $MD$ or its
heterogeneity take on. This means that, even if the condition in
@eq-ineq_short is violated, it is by no means guaranteed that an
attenuation correction procedure would actually reduce heterogeneity in
ES. This means that inference made from @eq-ineq_short is asymmetrical.
While we know, if the inequality holds, heterogeneity in corrected ES
will be larger than heterogeneity in uncorrected ES, we can't be certain
that the opposite is true if the inequality does not hold.

@@

Estimation of $R_1$ and $R_2$ requires estimates of meta-analytic mean
and heterogeneity of both the total score variance $\sigma^2_X$ and the
true score variance $\sigma^2_T$. Elsewhere, we proposed the method
Boot-Err (ref.) as an alternative method to test for and estimate the
differences in error score variance $\sigma^2_E$. Among other things, in
Beinhauer et al. (ref.), we deliver detailed instructions on how
estimates for mean and heterogeneity of total score variance, and the
score variance components can be derived. This means that Boot-Err can
be used to generate all estimates required for computation of the
metrics $R_1$ and $R_2$.

For the following analysis, Boot-Err is used. As Boot-Err makes use of
meta-analytic techniques and bootstrap sampling, we perform analyses
using the packages *metafor*, version 4.8-0, and *boot*, version 1.3-31,
in the statistical programming language *R*, version 4.4.2.

@@

Since $R_1$ is carrying the power of 3, it implies that the reduction in
mean score variance is affecting the ES heterogeneity more strongly than
the reduction in score variance heterogeneity. This is best described
through an example: We know what an acceptable average score reliability
looks like in the area of Psychology. Researchers largely target a score
reliability of at least .7, preferably .8. An average score reliability
of .7 implies that 70% of the variance observed in scores can actually
be attributed to true underlying differences. From our derived rule, we
know that, as long as $R_2$ is larger or equal to $.7^3 = .343$, the
attenuation correction procedure can only increase ES heterogeneity. If
we want to explain ES heterogeneity through score reliability, even
though only 30% of the variance observed in score can be attributed to
random measurement error, at least 65.7%, almost two thirds, of the
heterogeneity in score variance would need to be attributable to
differences in the error score variance. In other words, at most 34.3%
of heterogeneity in score variances may be due to actual differences in
how the latent variable is distributed (in terms of spread) across the
administration sites.

In the examples we re-analysed throughout this text, as well as the
majority of research studies done in the behavioural sciences, no proper
sampling techniques are employed. Most samples where these instruments
are applied come to be through convenience sampling, where students
self-select into studies. Without a valid sampling scheme, we
essentially have have no basis to assume anything about the true score
variance. Most importantly, however, we have no reason to assume that
the true score variance could potentially be homogeneous or stable
across administration sites. Instead, it is very likely that the
convenience sampling procedures produce samples that vary extremely
strongly in how they are put together, leading to large degrees of true
score variance heterogeneity. At the same time, score reliability is a
low-hanging fruit to optimise as most researchers are aware and trained
in assessing score reliability.

@@

Our results fit in well with recent work by Olsson-Collentine et al.
(2023). In a large simulation scheme, they demonstrate that differences
in score reliability across administrations typically deflate
heterogeneity in uncorrected correlations. In such cases, an attenuation
correction would increase correlation coefficient heterogeneity, as we
observe for standardized mean differences. Only as the true
heterogeneity in correlations grows larger would the correction
procedure reduce heterogeneity as implied in W&D. While correlations and
standardized effect sizes are not identical, the way score
(un)reliability affects these parameters is highly similar.

@@

The analytical arguments, detailed in Appendix X, leading to
@eq-ineq_short, rest on the use of the delta-method. @eq-ineq_init
contains parameters of the score standard deviations’ distributions,
while the $R_1$ and $R_2$ metrics were designed to discuss score
variances. The delta-method rests on a number of assumptions, at least
one of which is likely violated in our analysis: the variance of the
variable to be transformed needs to be small (ref.). This entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability.
Therefore, assuming that this assumption holds seems naive. In the
Appendix X, we explore the archival data, to understand in how far
employing the delta-method on meta-analyses of standard deviations lead
to biased results than the meta-analyses on score variances we performed
using the Boot-Err technique. The differences identified are rather
miniscule for the majority of phenomena, at the third decimal. Only for
the PSACR002_behav_init phenomenon we find larger differences -
therefore, conclusions for this phenomenon concerning the relation
between $R_1$ and $R_2$ need to be drawn with caution. Generally, as the
majority of phenomena did not display any bias in this assessment and
since the delta-method is not actually used to generate estimates, we
believe its use to be defensible nonetheless.

@@

The work done by Hunter and Schmidt (2014) was crucial in informing and
guiding methodology development in the field of meta-analyses.
Similarly, Wiernik and Dahlke (2020) raise a number of important points
that have been neglected in the application of meta-analytic research
over the last decades. We generally agree with their ideas that
underappreciated (differences in un-)reliability can introduce
substantial biases in meta-analytic estimates and tests in Psychology.
However, based on our findings, we believe that the issues created by
ignoring random sampling is much more dire and warrants immediate
attention. Only once we can be sure that the variance in true scores or
latent factors is actually meaningful for some population, we can turn
towards imprecise measurements, expecting correction procedures to
explain unwarranted heterogeneity. In the current state of Psychology,
meta-analysts should not expect that correcting ES for score reliability
will actually explain heterogeneity identified.

@@

Using the methdology underlying the Boot-Err approach (Beinhauer et al.
ref), we generate estimates of meta-analytic mean and heterogeneity in
log-transformed total score variance $\sigma^2_X$. To facilitate
interpretation and allow for computation of the metrics $R_1$ and $R_2$,
we need to back-transform those estimates to the original scale. Using
@eq-mu_sigmaX we can generate an estimate of meta-analytic mean score
variance on the scale of $\sigma^2_X$ and using @eq-tau_sigmaX we get an
estimate of its heterogeneity.

@@

In [equations @eq-mu_sigmaE] and [-@eq-tau_sigmaE] we generate estimates
of meta-analytic mean and heterogeneity of error score variance
$\sigma^2_E$. Just like for the total score variance, we make use of the
estimates of mean and heterogeneity from the meta-analysis on
log-transformed error score variance.

>>>>

Large-scale attempts of direct replications, using identical protocols,
such as the Many Labs studies or Registered Replication Reports [e.g.
@ManyLabs1; @RRRHart], for the first time allow researchers to estimate
heterogeneity undistorted by selection processes like publication bias.
In re-analyses of these studies, Olsson-Collentine et al. [-@olsson2020]
identify a positive correlation between a phenomenon’s effect size and
its heterogeneity. Similarly, van Erp et al. [-@vanErp2017] and Stanley
et al. [-@stanley2018] estimate strong degrees of heterogeneity across
conceptual replications in Psychology. In a separate re-analysis of
large scale direct replications, Renkewitz et al. (in preparation)
identify substantial heterogeneity in almost all projects where a
non-zero effect could be identified. This aligns with the correlation
found by Olsson-Collentine et al. [-@olsson2020].

@@

The ManyLabs projects were collaborative efforts to replicate several
psychological phenomena across different research sites, employing
identical protocols. From five published projects, the data for the
first three and the fifth Many Labs projects [@ManyLabs1; @ManyLabs2;
@ManyLabs3; @ManyLabs5] was harmonized and publicly stored in a
standardized format when the DRIPHT repository was set up [@dripht2024].
The Registered Replication Reports are similar collaborative efforts,
with the sole distinction that for each report a single phenomenon was
replicated several times. The Registered Reports 3-10 were added to the
DRIPHT repository, as they employed experimental designs with at least
two groups [@RRRHart; @RRRSripada; @RRRFinkel; @RRRStrack; @RRRRand;
@RRRDijksterhuis; @RRRSrull; @RRRMazar]. Lastly, some projects from the
Psychological Science Accelerator were added to the DRIPHT repository,
namely PSA006, PSACR001, PSACR002 and PSACR003 [@PSA006; @PSACR001;
@PSACR002; @PSACR003]. In contrast to Many Labs or Registered
Replication Projects, these collaborative efforts, while also
distributed across the globe, do not formally attempt to perform direct
replications, but focus on original research or conceptual replications.
However, as each project used the same protocol and participants are
distributed across different countries, a data structure similar to that
of a multi-site replication study emerges. What makes all these
collaborative efforts valuable for this manuscript is the fact that all
phenomena collected in the DRIPHT-repository are making use of two-group
designs and are therefore easily assessed using standardised mean
differences.

@@

For each replication of a phenomenon, Cohen’s $d$ is computed as an
estimate of raw ES, according to @eq-d, including its standard error
(@eq-SEd). Subsequently, a random-effects meta-analysis is performed
using metafor version 4.8-0 [@metafor]. To estimate meta-analytic mean
ES and heterogeneity we make use of the REML-estimator, as that
estimator tends to perform best across a range of conditions
[@hoenekopp2022]. To identify which phenomena pass criterion (3), using
a Wald-type significance test we assess for which phenomenon the
meta-analytic mean ES is significantly different from zero. For this
hypothesis test, as all other hypothesis tests in this manuscript, we
employ a one-tailed significance level of .05.

Additionally, estimates of score reliability are derived. While not
without criticism (references), Cronbach’s Alpha is used to estimate
score reliability. It is often noted that tau-equivalence is an
unrealistic assumption to hold against real-world data, implying that
other estimates of score reliability, such as McDonald’s Omega
[@mcdonald1999], Guttman’s Lambda 2 or 4 [@guttman1945], or the Greatest
Lower Bound [@tenBerge2004] might be better suited. However, all scales
employed in this project are scored by computing the simple mean or sum
of responses across items for each individual. Computing the mean or sum
of several items implicitly assumes tau-equivalence, as each item
contributes equally to the individual’s test score. Therefore,
Cronbach’s Alpha is the better choice to estimate score reliability, as
it avoids a mismatch in assumptions between how the score is computed
and how score reliability is estimated. The individual estimates of
Cohen’s d are corrected using these estimates of Cronbach's Alpha,
according to @eq-dc, including the corrected standard errors (@eq-SEdc).
Thereby, estimates of ES are computed, which are corrected for imperfect
reliability. Again, a random-effects meta-analysis is run using metafor,
generating an estimate of heterogeneity in corrected ES. Different
indicators of heterogeneity are readily available, such as $I^2$, $H^2$
or the coefficient of variation $CV$[@borenstein2009; @higgins2002].
However, the descriptions found in H&S and W&D discuss the absolute
amount of heterogeneity in terms of variance ($\tau^2$) or standard
deviation ($\tau$). These parameters discuss the variability of the
standardized ES in the population in terms of variance or standard
deviation. Therefore, we also make use of absolute heterogeneity here.

Subsequently, in order to assess whether ES heterogeneity was indeed
reduced by the attenuation correction procedure, the estimates of
heterogeneity in uncorrected ES and corrected ES are compared. Using
Cochran’s *Q*-test [@cochran1954], we assess whether the estimates of
heterogeneity are statistically significantly different from 0. Only for
those projects where we have sufficient confidence that the individual
estimates of heterogeneity are larger than zero can we actually begin to
interpret whether there was any change.

To identify whether a reduction in ES heterogeneity is indeed
accompanied by differences in score reliability, the estimates of score
reliability are assessed meta-analytically as well. To do so, a
Reliability Generalization Meta-Analysis is performed [@vacha1998;
@sanchez2013]. This entails that individual estimates of score
reliability are adequately transformed using Bonett’s transformation:
$T_{\hat{\rho}} = ln(1 - \hat{\rho})$. This transformation has
variance-stabilising properties and therefore allows for adequate
inferences concerning heterogeneity in score reliability [@bonett2002].
Cochran’s Q-test is used to identify statistically significant
heterogeneity in transformed score reliability. Estimates derived from a
Reliability Generalization Meta-Analysis using these transformations can
be back-transformed to the original score reliability scale
[@beinhauer2025]:

@@

We executed these steps for each phenomenon. The statistical programming
language R, Version 4.4.2 [@RCore] is used for all data manipulation and
statistical analysis. Graphics are constructed using the R package
*ggplot2*, version 3.5.1 [@ggplot2].

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($\delta_0$) and corrected ES
($\delta$), which are central to assess the research questions posed in
section *Attenuation Correction*. Most importantly, @tbl-tauES
demonstrates that for all but one of the phenomena, the attenuation
correction procedure barely has an impact on the extent of ES
heterogeneity. However, while the heterogeneity for the
Srull_Behaviour_Hostility phenomenon seems reduced by about a third of
its initial value, the ES heterogeneity for this phenomenon is rather
small to begin with. Using a significance test, we can not distinguish
this value confidently from zero. This implies that for the majority of
phenomena, differences in score reliability can not be made responsible
for heterogeneity in standardized effect sizes.

@@

In the re-analysis of archival data in the previous section we
demonstrated that, at least for the multi-site direct replication and
PSA-projects, that score reliability does not hold any noteworthy value
as a potential moderator explaining ES heterogeneity. Additionally, we
demonstrated that claims formulated in W&D and H&S regarding how score
reliability increases ES heterogeneity is false. We found a large number
of phenomena where, contrary to popular claims, correcting for score
reliability increases ES heterogeneity, instead of decreasing it.

@@

It might help to think about linear models, when discussing the
intuition that might underlie the idea, that controlling for score
reliability should reduce ES heterogeneity. Heterogeneity describes
unexplained variation in a parameter's true distribution, that we model
using meta-analytical techniques [@borenstein2009; @higgins2002]. In
that way, the idea of heterogeneity is closely related to residual
variance, as we observe it in a simple linear model. Both describe
variation that could not be explained by the predictive variables in the
case of the simple linear model, or not explained by incorporating the
Standard Errors in the case of a meta-analytic model.

@@

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs [-@hunter2004, p. XX]. @eq-HS implies that
heterogeneity in $\delta_0$ is essentially a function in which
heterogeneity and mean value of $\delta$ are weighted by mean score
reliability $\rho_{xx‘}$ and its heterogeneity.

@@

We adjusted @eq-HS using the notation used throughout this text. Here,
$\delta_0$, carrying a 0 in the index, refers to the uncorrected ES,
undistorted by sampling error but not corrected for measuring error. On
the other hand $\delta$, no index, refers to the corrected ES,
undistorted by sampling error and measurement error. Lastly,
$\rho_{XX'}$ refers to score reliability. Hunter & Schmidt
[-@hunter2004] claim that @eq-HS demonstrates that differences in score
reliability inflate ES heterogeneity, which means that an attenuation
correction procedure would have to reduce ES heterogeneity. However, it
can be easily verified that @eq-HS does not allow for such claims.
Depending on both the mean value $\mu_{\rho_{xx‘}}$ and the
heterogeneity $\tau^2_{\rho_{xx‘}}$ in score reliability, ES
heterogeneity might be increased or decreased by the differences in
score reliability.

Most likely, Hunter and Schmidt did not fully explore the implications
of @eq-HS, but followed the general intuition we lined out–"If a source
of variation is controlled for, the amount of unexplained variance
should be reduced." Wiernik and Dahlke [-@wiernik2020] refer to the work
by Hunter & Schmidt [-@wiernik2020], likely following their intuition.

In the re-analysis of archival data, we have demonstrated that this
intuition does not hold for the relationship of ES heterogeneity and
score reliability. Instead, we will propose an alternative way that
helps us discern, under which conditions we can expect an attenuation
correction procedure to lead to an increase in ES heterogeneity.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for a different analytical description of how differences in score
reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$. First order taylor approximation may be used to
generate estimates of mean and variance of the ratio variable
\[irrespective of distribution?\] [@vanKempen1999]. Here we use the
Taylor-estimator of the ratio’s variance to demonstrate how differences
in score reliability may affect heterogeneity in ES.

@@

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially describes the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_short states the following regarding the
change in mean value and heterogeneity due to the correction: ES
heterogeneity will always be reduced, as long as the remaining
heterogeneity in score variance is not substantially smaller than the
size of the remaining mean value.

@@

It is important to reiterate that the condition in @eq-ineq_short only
describes under which conditions the inequality is guaranteed to hold,
no matter what value other parameters like the mean value of $MD$ or its
heterogeneity take on. This means that, if the condition in
@eq-ineq_short is violated, it is by no means guaranteed that an
attenuation correction procedure would actually reduce heterogeneity in
ES. Thereby, inference made from @eq-ineq_short is asymmetrical. While
we know, if the inequality holds, heterogeneity in corrected ES will be
larger than heterogeneity in uncorrected ES, we can't be certain that
the opposite is true if the inequality does not hold.

@@

Estimation of $R_1$ and $R_2$ requires estimates of meta-analytic mean
and heterogeneity of both the total score variance $\sigma^2_X$ and the
true score variance $\sigma^2_T$. Elsewhere, we proposed the method
Boot-Err (ref.) as an alternative method to test for and estimate the
differences in error score variance $\sigma^2_E$. Among other things, in
Beinhauer et al. [-@beinhauer2025] we deliver detailed instructions on
how estimates for mean and heterogeneity of total score variance, and
the score variance components can be derived. This means that Boot-Err
can be used to generate all estimates required for computation of the
metrics $R_1$ and $R_2$.

For the following analysis, Boot-Err is used. As Boot-Err makes use of
meta-analytic techniques and bootstrap sampling, we perform analyses
using the packages *metafor*, version 4.8-0 [@metafor], and *boot*,
version 1.3-31 [@boot], in the statistical programming language *R*,
version 4.4.2 [@RCore].

@@

Since $R_1$ is carrying the exponent of 3, it implies that the reduction
in mean score variance is affecting the ES heterogeneity more strongly
than the reduction in score variance heterogeneity. This is best
described through an example: We know what an acceptable average score
reliability looks like in the area of Psychology. Researchers largely
target a score reliability of at least .7, preferably .8. An average
score reliability of .7 implies that 70% of the variance observed in
scores can actually be attributed to true underlying differences. From
our derived rule, we know that, as long as $R_2$ is larger or equal to
$.7^3 = .343$, the attenuation correction procedure can only increase ES
heterogeneity. If we want to explain ES heterogeneity through score
reliability, even though only 30% of the variance observed in score can
be attributed to random measurement error, at least 65.7%, almost two
thirds, of the heterogeneity in score variance would need to be
attributable to differences in the error score variance. In other words,
at most 34.3% of heterogeneity in score variances may be due to actual
differences in how the latent variable is distributed (in terms of
spread) across the administration sites.

In the examples we re-analysed throughout this text, as well as the
majority of research studies done in the behavioural sciences
[@hultsch2002; @bornstein2013], no proper sampling techniques are
employed. Most samples where these instruments are applied come to be
through convenience sampling, where students self-select into studies.
Without a valid sampling scheme, we essentially have have no basis to
assume anything about the true score variance [@hultsch2002]. Most
importantly, however, we have no reason to assume that the true score
variance could potentially be homogeneous or stable across
administration sites. Instead, it is very likely that the convenience
sampling procedures produce samples that vary extremely strongly in how
they are put together, leading to large degrees of true score variance
heterogeneity. At the same time, score reliability is a low-hanging
fruit to optimise as most researchers are aware and trained in assessing
score reliability [@hussey2025].

@@

Our results fit in well with recent work by Olsson-Collentine et al.
[@olsson2023]. In a large simulation scheme, they demonstrate that
differences in score reliability across administrations typically
deflate heterogeneity in uncorrected correlations. In such cases, an
attenuation correction would increase correlation coefficient
heterogeneity, as we observe for standardized mean differences. Only as
the true heterogeneity in correlations grows larger would the correction
procedure reduce heterogeneity as implied in W&D. While correlations and
standardized effect sizes are not identical, the way score
(un)reliability affects these parameters is highly similar.

@@

The analytical arguments, detailed in *Appendix A*, leading to
@eq-ineq_short, rest on the use of the delta-method. @eq-ineq_init
contains parameters of the score standard deviations’ distributions,
while the $R_1$ and $R_2$ metrics were designed to discuss score
variances. The delta-method rests on a number of assumptions, at least
one of which is likely violated in our analysis: the variance of the
variable to be transformed needs to be small [@cramer1999]. This entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability.
Therefore, assuming that this assumption holds seems naive. In the
*Appendix C*, we explore the archival data, to understand in how far
employing the delta-method on meta-analyses of standard deviations lead
to biased results than the meta-analyses on score variances we performed
using the Boot-Err technique. The differences identified are rather
miniscule for the majority of phenomena, at the third decimal. Only for
the PSACR002_behav_init phenomenon we find larger differences -
therefore, conclusions for this phenomenon concerning the relation
between $R_1$ and $R_2$ need to be drawn with caution. Generally, as the
majority of phenomena did not display any bias in this assessment and
since the delta-method is not actually used to generate estimates, we
believe its use to be defensible nonetheless.

@@

The work done by Hunter and Schmidt [-@hunter2004] was crucial in
informing and guiding methodology development in the field of
meta-analyses. Similarly, Wiernik and Dahlke [-@wiernik2020] raise a
number of important points that go far beyond attenuation correction and
have been neglected in the application of meta-analytic research over
the last decades. We generally agree with their ideas that
underappreciated (differences in un-)reliability can introduce
substantial biases in meta-analytic estimates and tests in Psychology.
However, based on our findings, we believe that the issues created by
ignoring random sampling is much more dire and warrants immediate
attention. Only once we can be sure that the variance in true scores or
latent factors is actually meaningful for some population, we can turn
towards imprecise measurements, expecting correction procedures to
explain unwarranted heterogeneity. In the current state of Psychology,
meta-analysts should not expect that correcting ES for score reliability
will actually explain heterogeneity identified.

@@

Using the methoodlogy underlying the Boot-Err approach [@beinhauer2025],
we generate estimates of meta-analytic mean and heterogeneity in
log-transformed total score variance $\sigma^2_X$. To facilitate
interpretation and allow for computation of the metrics $R_1$ and $R_2$,
we need to back-transform those estimates to the original scale. Using
@eq-mu_sigmaX we can generate an estimate of meta-analytic mean score
variance on the scale of $\sigma^2_X$ and using @eq-tau_sigmaX we get an
estimate of its heterogeneity.

@@

In [Equations @eq-mu_sigmaE] and [-@eq-tau_sigmaE] we generate estimates
of meta-analytic mean and heterogeneity of error score variance
$\sigma^2_E$. Just like for the total score variance, we make use of the
estimates of mean and heterogeneity from the meta-analysis on
log-transformed error score variance.
